First paper - Draft

### 2017-07-05 paper/paper.Rnw
All graphs and tables visible in tex. Rearranging and recommenting

\usepackage{geometry}
 \geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
 }
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[round, colon]{natbib}
\usepackage[colorlinks]{hyperref}
\AtBeginDocument{%
  \hypersetup{
    citecolor=blue,
    linkcolor=blue,   
    urlcolor=blue}}

@@

<<default_conf>>=
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)
  library(xtable)

@@

% Figure out how to ouput data frames as latex tables.
<<sample_size, results = 'tex'>>=
# Get sample counts for each dummy
sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {

@@

sample_tables_topbottom <- sample_size_calc(adapted_year_data, c(.1, .9), selected = TRUE, countries)
sample_tables_topmid <- sample_size_calc(adapted_year_data, c(.5, .9), selected = TRUE, countries)
sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), selected = TRUE, countries)

summary_table <-
  left_join(summary_data, sample_tables_topbottom) %>%
  filter(name %in% c("2000", "2015")) %>%
  arrange(country, name)

@@

print(table_coming,
      add.to.row = addtorow,
      include.colnames = FALSE,
      include.rownames = FALSE,
      hline.after = NULL,
      caption.placement = "top",
      caption = "SES sample size and ISCED composition")
@

@@

<<correlation_incomeineq, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
complete_data_topbottom %>%
  mutate(wave = as.character(wave)) %>%
  left_join(inequalityintsvy::economic_inequality, by = c("wave" = "year", "country")) %>%
  filter(indicators == "GINI") %>%
  group_by(country) %>%
  summarize(avg_diff = mean(difference, na.rm = T),
            avg_value = mean(value, na.rm = T)) %>%
  filter(avg_value < 8) %>%
  ggplot(aes(avg_value, avg_diff)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), linetype = "longdash", se = F)
@

@@

<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# 90/10 gaps acros countries
diff_increase_fun <- function(df) {

@@

complete_data_topbottom %>%
  filter(country %in% countries) %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 3), linetype = "longdash",
            colour = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  geom_line(alpha = 0.4) +
  geom_point(size = 0.5, alpha = 0.4) +
  coord_cartesian(ylim = c(0, 3)) +
  facet_wrap(~ country)
@

We start by look at the achievement for some countries. We plot both mathematics and reading gaps for each country and also a cubic trend spline for both test scores pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden. In mathematics, France increase the gap by roughly \Sexpr{diff_data$math[["France"]]}. This pattern happens similarly for literacy with an increase of \Sexpr{diff_data$read[["France"]]}. For such a short period of time, the magnitud of the increase is reasonably big.

@@

However, we don't know what happens over time. But Ballarino and Bernardi (2016) did study the relationship of social origin effects and found that they increased very much in France in the last decades.

@@

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation. The actual difference is of about \Sexpr{diff_data$math[["Sweden"]]}. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

Something reassuring is that mathematics and reading follow basically the same trend across all countries. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving why the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

<<>>=
# You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
complete_data_topbottom %>%
  filter(country %in% countries) %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  select(wave, country, matches("*._math$")) %>%
  gather(ses, gap_size, matches("^\\d_mean_math$")) %>%
  separate(ses, c("ses", "delete"), sep = 2) %>%
  mutate(ses = gsub("_", "", ses)) %>%
  filter(complete.cases(.)) %>%
  # gather(lower_bound, lower, matches("\\d_lower")) %>%
  # gather(upper_bound, upper, matches("\\d_upper"))
  ggplot(aes(as.factor(wave), gap_size, group = ses, colour = ses)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_line(alpha = 0.4) +
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 1), linetype = "longdash",
            colour = "black") +
  # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
  # geom_hline(yintercept = 0, linetype = "longdash") +
  # coord_cartesian(ylim = c(0, 3)) +
  facet_wrap(~ country)
@

<<>>=

complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "math", !is.na(country)) %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 12) +
  viridis::scale_fill_viridis(option = "B") +
  scale_y_continuous(expand = c(0, 0), lim = c(0, 3)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
      axis.text.x = element_text(angle = 90, hjust = 1))

@
The top line belongs to the high SES group whereas the bottom line belongs to the low SES group. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

@@

Ths source of this gap can be deconstructed into two broader gaps. We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group. Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap. For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th? Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier. Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous plots \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.

@@

<<graphing_allgaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

# Comparing all gaps across countries
complete_data_topbottom %>%
  bind_rows(complete_data_topmid) %>%
  bind_rows(complete_data_midbottom) %>%
  filter(country %in% c("United States", "Denmark", "France", "Germany")) %>%
  mutate(type = factor(type,
                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
                       ordered = TRUE)) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(-0.5, 3)) +
  facet_grid(country ~ type)
@

@@

<<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# Graphing how the top/bottom are evolving over time instead of absolute difference

complete_data_topbottom %>%
  bind_rows(complete_data_topmid) %>%
  bind_rows(complete_data_midbottom) %>%
  filter(country %in% c("United States", "Denmark", "France", "Germany"), type_test == "math") %>%
  select(wave, country, type_test, type, contains("math")) %>%
  mutate(type = factor(type,
                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
                       ordered = TRUE)) %>%
  gather(score, value, -(wave:type)) %>%
  separate(score, c("ses", "score"), sep = 2) %>%
  spread(score, value) %>%
  ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses)) +
  geom_errorbar(aes(ymin = lower_math, ymax = upper_math), width = 0.1) +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(-0.5, 3)) +
  facet_grid(country ~ type)

# Increase:  
# Sweden - steady increase in both tests
# Austria - increase in math - slight increase in read
# Finland - very sharp increase in both
# France - very sharp increase in both
# Netherlands - sharp increase in both

# Decrease:
# US - decrease in both tests
# Chile - decrease in both tests

# No change:
# Canada - stable red - increase math
# UK - slight decrease red - stable math
# Belgium - no change
# Czech republic - no change
# Denmark no change
# Germany - no change
# ITaly - no change
# Japan -  no change
# Norway - no change
# Poland - no change
# Spain - no change
@

For all country-years pooled we see a very strong correlation but contrary to what we would expect. We find that countries with high performance have also a very low level of inequality. In fact, even if we separate it by year.

<<>>=
data_to_plot %>%
  ggplot(aes(scale(Mean), difference)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ wave)
@

<<>>=
data_to_plot %>%
  filter(country %in% countries) %>%
  ggplot(aes(as.character(wave), difference, group = country)) +
  geom_point(colour = "blue") +
  geom_line(colour = "blue") +
  geom_point(aes(y = scale(Mean)), colour = "red") +
  geom_line(aes(y = scale(Mean)), colour = "red") +
  facet_wrap(~ country)
@

In fact, we see a strong correspondence. In countries like Germany and France we see quite a functional form between both inequality and performance. This pattern is also seen in in the UK, Austria, Australia and New Zealand. Poland and Italy are quite interesting cases because it shows how inequality has been decreasing that much whereas average performance has been improving notably. On another hand, only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.

<<>>=
data_to_plot %>%
  semi_join(data_to_plot %>% group_by(country) %>% count() %>% filter(n > 3), by = "country") %>%
  group_by(country) %>%
  summarize(cor = cor(scale(Mean), difference, use = 'complete.obs')) %>%
  ggplot(aes(reorder(country, cor), cor, fill = cor > 0)) + geom_col() + coord_flip()
@

@@

<<rate_change_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.05, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@

In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing. We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

<<graphing_perc_increase, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
perc_graph <- function(df, title) {
  df %>%
  enframe(name = "country") %>%
  unnest(value) %>%
  filter(country %in% countries) %>%
  select(-country1, -years_available) %>%
  mutate(diff_95_lower = perc_diff - 2*sd_year,
         diff_95_upper = perc_diff + 2*sd_year) %>%
  setNames(c("Country", "Type of test", "Average % difference", "Average SD",
             "Lower 50% bound", "Upper 50% bound", "Lower 95% bound", "Upper 95% bound")) %>%
  filter(`Type of test` == "math") %>%
  arrange(`Average % difference`) %>%
  mutate(Country = ordered(forcats::as_factor(Country)),
         Country_num = as.numeric(Country)) %>%
  ggplot(aes(Country, `Average % difference`)) +
  geom_point() +
  geom_ribbon(aes(x = Country_num, ymin = `Lower 95% bound`, ymax = `Upper 95% bound`), alpha = 0.5) +
  geom_ribbon(aes(x = Country_num, ymin = `Lower 50% bound`, ymax = `Upper 50% bound`), alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  scale_y_continuous(breaks = seq(-160, 160, 40)) +
  coord_cartesian(ylim = c(-160, 160)) +
  ggtitle(title) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

m1 <- perc_graph(top_bottom_perc, "90/10")
m2 <- perc_graph(top_mid_perc, "90/50")
m3 <- perc_graph(mid_bottom_perc, "50/10")

gridExtra::grid.arrange(m1, m2, m3, nrow = 1)
@

>>>>

\usepackage{geometry}
 \geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
 }
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[round, colon]{natbib}
\usepackage[colorlinks]{hyperref}
\usepackage{booktabs}
\AtBeginDocument{%
  \hypersetup{
    citecolor=blue,
    linkcolor=blue,   
    urlcolor=blue}}

@@

<<default_conf>>=
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)
  library(xtable)
  library(ggthemes)

@@

% Figure out how to ouput data frames as latex tables.
<<sample_size>>=
# Get sample counts for each dummy
sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {

@@

sample_tables_topbottom <- sample_size_calc(adapted_year_data, c(.1, .9), selected = TRUE, countries)
sample_tables_topmid <- sample_size_calc(adapted_year_data, c(.5, .9), selected = TRUE, countries)
sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), selected = TRUE, countries)
@

<<table_escs, include = TRUE>>=
summary_table <-
  left_join(summary_data, sample_tables_topbottom) %>%
  filter(name %in% c("2000", "2015")) %>%
  arrange(country, name)

@@

print(table_coming,
      scalebox = '0.75',
      floating = FALSE,
      add.to.row = addtorow,
      include.colnames = FALSE,
      include.rownames = FALSE,
      hline.after = nrow(table_coming),
      caption.placement = "top",
      caption = "SES sample size and ISCED composition")
@

@@

<<correlation_incomeineq, include = F, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
complete_data_topbottom %>%
  mutate(wave = as.character(wave)) %>%
  left_join(inequalityintsvy::economic_inequality, by = c("wave" = "year", "country")) %>%
  filter(indicators == "GINI") %>%
  group_by(country) %>%
  summarize(avg_diff = mean(difference, na.rm = T),
            avg_value = mean(value, na.rm = T)) %>%
  filter(avg_value < 8) %>%
  ggplot(aes(avg_value, avg_diff)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), linetype = "longdash", se = F)
@

@@

\begin{figure}
\begin{center}
<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# 90/10 gaps acros countries
diff_increase_fun <- function(df) {

@@

complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "read", !is.na(country)) %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 5) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")
@
\end{center}
\end{figure}

We start by look at the achievement for some countries. We plot only mathematics for each country and also a quadratic trend spline for both test scores pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden. In mathematics, France increase the gap by roughly \Sexpr{diff_data$math[["France"]]}. This pattern happens similarly for literacy with an increase of \Sexpr{diff_data$read[["France"]]}. For such a short period of time, the magnitud of the increase is reasonably big.

@@

However, we don't know what happens over time. But \citet{bernardi2016} did study the relationship of social origin effects and found that they increased very much in France in the last decades.

@@

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015. The actual difference is of about \Sexpr{diff_data$math[["Sweden"]]}. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

The plot below takes a more direct approach and looks at the percentage change from the first and last timepoint available. Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.

\begin{figure}
\begin{center}
<<graphing_perc_increase, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
perc_graph <- function(df, test, title, subtitle = NULL) {
  df %>%
  enframe(name = "country") %>%
  unnest(value) %>%
  filter(country %in% countries) %>%
  select(-country1, -years_available) %>%
  mutate(diff_95_lower = perc_diff - 2*sd_year,
         diff_95_upper = perc_diff + 2*sd_year) %>%
  setNames(c("Country", "Type of test", "Average % difference", "Average SD",
             "Lower 50% bound", "Upper 50% bound", "Lower 95% bound", "Upper 95% bound")) %>%
  filter(`Type of test` == test) %>%
  arrange(`Average % difference`) %>%
  mutate(Country = ordered(forcats::as_factor(Country)),
         Country_num = as.numeric(Country)) %>%
  ggplot(aes(Country, `Average % difference`, fill = `Average % difference` > 0)) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  geom_point() +
  geom_ribbon(aes(x = Country_num, ymin = `Lower 95% bound`, ymax = `Upper 95% bound`), alpha = 0.3) +
  geom_ribbon(aes(x = Country_num, ymin = `Lower 50% bound`, ymax = `Upper 50% bound`), alpha = 0.2) +
  geom_linerange(aes(ymin = 0, ymax = `Average % difference`,
                     colour = `Average % difference` > 0), alpha = 0.4) +
  scale_y_continuous(name = "Avg % increase/decrease",breaks = seq(-160, 160, 40)) +
  scale_fill_discrete(guide = FALSE) +
  scale_colour_discrete(guide = FALSE) +
  coord_cartesian(ylim = c(-160, 160)) +
  ggtitle(title, subtitle) +
  ggthemes::theme_few() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

perc_graph(top_bottom_perc, "math", "90/10 achievement gap", "Percentage change from 2000 to 2015")
# m2 <- perc_graph(top_mid_perc, "90/50")
# m3 <- perc_graph(mid_bottom_perc, "50/10")

# gridExtra::grid.arrange(m1, m2, m3, nrow = 1)
@
\end{center}
\end{figure}

Generally speaking, we see that most countries increased their achievement gap. France had an average increase of about 80\% since 2000, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.

<<evolution_reading, eval = F>>=
# See below that reading graph is upon request. This is the one.
complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "read", !is.na(country)) %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 5) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  coord_fixed() +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")

perc_graph(top_bottom_perc, "read", "90/10 achievement gap", "Percentage change from 2000 to 2015")

Something reassuring is that mathematics and reading follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving why the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

\begin{figure}
\begin{center}
<<rate_increase/decrease, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
complete_data_topbottom %>%
  filter(country %in% countries) %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  select(wave, country, matches("*._math$")) %>%
  gather(ses, gap_size, matches("^\\d_mean_math$")) %>%
  separate(ses, c("ses", "delete"), sep = 2) %>%
  mutate(ses = gsub("_", "", ses)) %>%
  filter(complete.cases(.)) %>%
  # gather(lower_bound, lower, matches("\\d_lower")) %>%
  # gather(upper_bound, upper, matches("\\d_upper"))
  ggplot(aes(as.factor(wave), gap_size, group = ses, colour = ses)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_line(alpha = 0.4) +
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 1), linetype = "longdash",
            colour = "black") +
  scale_y_continuous(name = "Standardized test scores (mean 0)") +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
  ggtitle("Evolution of the achievement gap by top/bottom groups") +
  # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
  # geom_hline(yintercept = 0, linetype = "longdash") +
  # coord_cartesian(ylim = c(0, 3)) +
  facet_wrap(~ country, ncol = 5) +
  theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        legend.position = "bottom")
@
\end{center}
\end{figure}

The top line belongs to the high SES group whereas the bottom line belongs to the low SES group. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

@@

Ths source of this gap can be deconstructed into two broader gaps. We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group. Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap. For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th? Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier? Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.

@@

\begin{figure}
\begin{center}
<<graphing_allgaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

# Comparing all gaps across countries
complete_data_topbottom %>%
  bind_rows(complete_data_topmid) %>%
  bind_rows(complete_data_midbottom) %>%
  filter(country %in% c("United States", "Denmark", "France", "Germany")) %>%
  mutate(type = factor(type,
                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
                       ordered = TRUE)) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_line(alpha = 0.4) +
  geom_point(size = 0.5, alpha = 0.4) +
  coord_cartesian(ylim = c(0, 3)) +
  scale_y_continuous(name = "Standardized test scores (mean 0)") +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
  ggtitle("Comparison of the achievement gap for three socio-economic groups") +
  facet_grid(country ~ type) +
  theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        legend.position = "bottom")
@
\end{center}
\end{figure}

@@

In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing.

@@

\begin{figure}
\begin{center}
<<rate_change_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.05, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@
\end{center}
\end{figure}

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

\begin{figure}
\begin{center}
<<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# Graphing how the top/bottom are evolving over time instead of absolute difference

complete_data_topbottom %>%
  bind_rows(complete_data_topmid) %>%
  bind_rows(complete_data_midbottom) %>%
  filter(country %in% c("United States", "Denmark", "France", "Germany"), type_test == "math") %>%
  select(wave, country, type_test, type, contains("math")) %>%
  mutate(type = factor(type,
                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
                       ordered = TRUE)) %>%
  gather(score, value, -(wave:type)) %>%
  separate(score, c("ses", "score"), sep = 2) %>%
  spread(score, value) %>%
  ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses)) +
  geom_errorbar(aes(ymin = lower_math, ymax = upper_math), width = 0.1) +
  geom_line(alpha = 0.4) +
  geom_point(size = 0.5, alpha = 0.4) +
  coord_cartesian(ylim = c(-1, 2)) +
  scale_y_continuous(name = "Standardized test scores (mean 0)") +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
  ggtitle("Evolution of the top/bottom groups by different gaps") +
  facet_grid(country ~ type) +
  theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        legend.position = "bottom")
@
\end{center}
\end{figure}

<<correlation_performance_inequality>>=
correlation <-
  data_to_plot %>% group_by(wave) %>%
  summarize(cor = cor(difference, Mean, use = 'complete.obs')) %>%
  pull(cor)
@

\begin{figure}
\begin{center}
<<avgperformance_inequality, include = T, out.height = '5in', out.width = '8.5in', fig.align = 'center'>>=
data_to_plot %>%
  ggplot(aes(scale(Mean), difference)) +
  geom_smooth(method = "lm") +
  geom_point() +
  scale_y_continuous(name = "90/10 achievement gap") +
  scale_x_continuous(name = "Average performance standardized") +
  ggtitle("Relationship between average performance and achievement gap") +
  facet_wrap(~ wave, nrow = 2) +
  theme_few()
@
\end{center}
\end{figure}

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between \Sexpr{max(correlation)} and \Sexpr{min(correlation)}. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries.
